---
title: "Final Project_Digital Strategies in Social Science"
author: "Mishu Dhar"
format: pdf
editor: visual
---

## Importing Libraries

```{r}
#| echo: false
#| message: false
#| warning: false
#| include: false

library(tidyverse)
library(ggplot2)
library(dplyr)
library(tidyr)
library(factoextra)
library(FactoMineR)
```

# Importing dataset

```{r}
#| echo: false
#| message: false
#| warning: false


socialism <- read.csv("/Users/mishudhar/Desktop/DSSC/Socialism/Socialism.csv")
# checking few rows
head(socialism)
```

```{r}
# column names
#colnames(socialism)
colSums(is.na(socialism))
```

```{r}
dim(socialism)
```

```{r}
# filtering for selected varoables
socialism_filtered <- socialism |>
  select(ccode, country, exp.environment, exp.total, exp.consumption, exp.housing,
         exp.health, exp.recreation, exp.education, soc.health, revenues, tax.supra.tot,
         exp.socprot, population, revenues)
colnames(socialism_filtered)
```

```{r}
# first few rows
head(socialism_filtered)
```

```{r}

# Rename the columns in the dataframe
socialism_filtered <- socialism_filtered %>%
  rename(
    population = population,
    exp_environment = exp.environment,
    exp_total = exp.total,
    exp_consumption = exp.consumption,
    exp_housing = exp.housing,
    exp_health = exp.health,
    exp_recreation = exp.recreation,
    exp_education = exp.education,
    socialSupport_health = soc.health,
    state_tax_total = tax.supra.tot,
    social_protection = exp.socprot,
    revenues = revenues
    
  )

# Calculate the missing values
missing_values <- colSums(is.na(socialism_filtered))

# Create a data frame of missing values counts
missing_values_df <- data.frame(
  column = names(missing_values),
  missing_count = as.numeric(missing_values)
)

# Filter to only include columns with missing values
missing_values_df <- missing_values_df[missing_values_df$missing_count > 0, ]

# Create the bar plot
ggplot(missing_values_df, aes(x = reorder(column, -missing_count), y = missing_count)) +
  geom_bar(stat = "identity", fill = "skyblue") +
  coord_flip() +  # Flip coordinates for better readability
  labs(title = "Missing Values Count by Column", x = "Columns", y = "Missing Values Count") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```

```{r}
# checking for missing values
colSums(is.na(socialism_filtered))
```

```{r}
# Filter the rows where the 'country' column is missing
missing_country_rows <- socialism_filtered %>% 
  filter(is.na(country))

# Display the rows
print(missing_country_rows)

```

```{r}


# Replace the missing 'country' name where 'ccode' is 255 with "Tanzania"
socialism_filtered <- socialism_filtered %>%
  mutate(country = ifelse(ccode == 255 & is.na(country), "Tanzania", country))

# Drop the other rows where 'country' is missing
socialism_filtered <- socialism_filtered %>%
  filter(!is.na(country))

# Verify the changes
print(socialism_filtered %>% filter(ccode == 255))
print(socialism_filtered %>% filter(is.na(country)))

colSums(is.na(socialism_filtered))
```

```{r}
# Filter the rows where the 'country' column is missing
missing_country_rows <- socialism_filtered %>% 
  filter(is.na(population))

# Display the rows
print(missing_country_rows)
```

```{r}
# Population data as whole numbers
population_data <- data.frame(
  ccode = c(344, 352, 355, 360, 365, 367, 368, 560, 710, 750, 850),
  population = c(4050000, 1210000, 6930000, 19050000, 145930000, 1870000, 2800000, 59310000, 1439323776, 1380004385, 273523615)
)

# Merge population data with the original dataframe
socialism_filtered <- socialism_filtered %>%
  left_join(population_data, by = "ccode") %>%
  mutate(population = coalesce(population.x, population.y)) %>%
  select(-population.x, -population.y)

```

```{r}
# Impute missing values by the values of the previous rows (observation)
socialism_df <- socialism_filtered %>%
  fill(everything(), .direction = "down")


colSums(is.na(socialism_df))
```

```{r}
# Calculate the mean of 'state_tax_total' excluding missing values
mean_state_tax_total <- mean(socialism_df$state_tax_total, na.rm = TRUE)

# Impute missing values with the mean for the specified rows
socialism_df <- socialism_df %>%
  mutate(state_tax_total = ifelse(ccode %in% c(2, 20, 70, 155) & is.na(state_tax_total), 
                                  mean_state_tax_total, 
                                  state_tax_total))

# Verify changes
print(socialism_df %>% filter(ccode %in% c(2, 20, 70, 155)))

```

```{r}
# socialism_df <- socil
# dropig the country code column
socialism_df <- socialism_df |>
  select(-ccode)

```

```{r}
# for future needs 
socialism_final <- socialism_df
```

```{r}
# Set the 'country' column as row names and remove it from the dataframe
row.names(socialism_df) <- socialism_df$country
socialism_df <- socialism_df %>%
  select(-country)

# Verify the column names
colnames(socialism_df)

```

```{r}
# check the data
head(socialism_df)

```

# PCA Principal Component Analysis

```{r}
# correaltion and scatterplot
library("GGally")

ggpairs(socialism_df)
```

```{r}
#| echo: false
#| message: false
#| warning: false


# Load necessary libraries
library(GGally)
library(ggplot2)

# Adjust the plot settings for better readability
ggpairs_plot <- ggpairs(
  socialism_df,
  title = "Correlation and Scatterplot Matrix",
  upper = list(continuous = wrap("cor", size = 4)),
  lower = list(continuous = wrap("points", size = 0.5)),
  diag = list(continuous = wrap("barDiag", size = 3))
) +
  theme_bw() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, size = 10),
    axis.text.y = element_text(size = 10),
    plot.title = element_text(hjust = 0.5, size = 14)
  )

# Print the plot
print(ggpairs_plot)

# Save the plot as an image file
ggsave("ggpairs_plot.png", plot = ggpairs_plot, width = 12, height = 12, dpi = 300)

```

```{r}
# principal component analysis
pc <- prcomp(socialism_df, center = TRUE, scale = TRUE)
# observing the summary
summary(pc)
```

First 8 principle components explain the 96 % of the variance of the data.

```{r}
# checking the attributes
attributes(pc)
```

```{r}
# loadings of the principal components
print(pc)
```

We can see that the variables have contributed in different principal components in various ways. In some PC's they have contributed very strongly positive and in some PC's they have contributed strongly negatively.

```{r}
# scree plöt
plot(pc)
```

```{r}
# scree plot line
screeplot(x = pc, type = "line", main  = "Elbow Method")
```

```{r}
# Load necessary library
library(factoextra)


# Plotting eigenvalues with a title
fviz_eig(pc) + 
  ggtitle("Eigenvalues of Principal Components") +
  theme_minimal()

```

If we take 8 PC's we will loss the 4 % of the data

```{r}
# using biplot
biplot(pc, cex = 0.6)
```

```{r}
# using autoplot
library(ggfortify)
autoplot(pc, scale = 0)
```

```{r}
# enhancing the plot
autoplot(pc, data = socialism_df,
         label = TRUE,
         label.size = 3,
         loadings = TRUE, color = "blue",
         loadings.label = TRUE,
         loadings.label.size = 3,
         repel = TRUE)
```

```{r}
# visualizing the individuals aka countries
fviz_pca_ind(pc,
             col.ind = "cos2",
             gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07"),
             repel = TRUE)
```

```{r}
# visualizing the variables
fviz_pca_var(pc,
             col.var = "contrib",
             gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07"),
             repel = TRUE)
```

```{r}
# plotting both variables and individuals
fviz_pca_biplot(pc, repel = TRUE,
                col.var = "black",
                col.ind = "cos2")
```

# Applying PCA to clustering

```{r}
# pca data
pca_df <- data.frame(pc$x[, 1:8])
head(pca_df)
```

```{r}
# determining number of clusters
fviz_nbclust(pca_df,
             FUNcluster = kmeans,
             method = "wss")
```

The elbow plot suggests that the optimal number of clusters is 5. This is identified at the point where the within-cluster sum of squares (WSS) starts to decrease at a slower rate. Beyond this point, additional clusters contribute less to the reduction in WSS, indicating diminishing returns in terms of improved clustering performance. Hence, selecting 5 clusters provides a balance between minimizing WSS and avoiding overfitting.

```{r}

optimal_clusters <- 5

# Perform K-means clustering
set.seed(123) # for reproducibility
kmeans_socialism <- kmeans(pc$x[, 1:8], centers = optimal_clusters, nstart = 25)

# Visualize clusters using PCA results
fviz_pca_ind(pc,
             geom.ind = "point", # show points only (nbut not "text")
             col.ind = as.factor(kmeans_socialism$cluster), # color by clusters
             palette = "jco",
             addEllipses = TRUE, # Concentration ellipses
             legend.title = "Clusters",
             repel = TRUE) # Avoid text overlapping (if text labels are shown)


```

```{r}
# Load necessary libraries
library(factoextra)
library(ggplot2)
library(ggrepel)

# Perform PCA
#pc <- prcomp(socialism_df, center = TRUE, scale. = TRUE)

# Perform K-means clustering
set.seed(123) # for reproducibility
optimal_clusters <- 3
kmeans_socialism <- kmeans(pc$x[, 1:8], centers = optimal_clusters, nstart = 25)

# Create a data frame for visualization
pca_data <- data.frame(pc$x[, 1:2], cluster = as.factor(kmeans_socialism$cluster), country = rownames(socialism_df))

# Visualize clusters with country labels
ggplot(pca_data, aes(x = PC1, y = PC2, color = cluster, label = country)) +
  geom_point(size = 2) +
  geom_text_repel(size = 3, max.overlaps = 20) +
  labs(title = "PCA Cluster Plot with Country Labels", x = "Principal Component 1", y = "Principal Component 2") +
  theme_minimal() +
  theme(legend.title = element_text(size = 10), legend.text = element_text(size = 8))

```

```{r}
# Load necessary libraries
library(factoextra)
library(ggplot2)
library(ggrepel)


# Perform K-means clustering
set.seed(123) # for reproducibility
optimal_clusters <- 5
kmeans_socialism <- kmeans(pc$x[, 1:8], centers = optimal_clusters, nstart = 25)

# Create a data frame for visualization
pca_data <- data.frame(pc$x[, 1:2], cluster = as.factor(kmeans_socialism$cluster), country = rownames(socialism_df))

# Visualize clusters with country labels and ellipses
ggplot(pca_data, aes(x = PC1, y = PC2, color = cluster, label = country)) +
  geom_point(size = 2) +
  geom_text_repel(size = 3, max.overlaps = 20) +
  stat_ellipse(aes(group = cluster), type = "norm", level = 0.95) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey") +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey") +
  labs(title = "PCA Cluster Plot with Country Labels and Ellipses",
       x = "Principal Component 1",
       y = "Principal Component 2") +
  theme_minimal() +
  theme(legend.title = element_text(size = 10), legend.text = element_text(size = 8))

```

```{r}
# Load necessary libraries
library(factoextra)
library(ggplot2)
library(ggrepel)

# Perform K-means clustering
set.seed(123) # for reproducibility
optimal_clusters <- 3  # Change this to 5 clusters
kmeans_socialism <- kmeans(pc$x[, 1:8], centers = optimal_clusters, nstart = 25)

# Create a data frame for visualization
pca_data <- data.frame(pc$x[, 1:2], cluster = as.factor(kmeans_socialism$cluster), country = rownames(socialism_df))

# Visualize clusters with country labels and ellipses
ggplot(pca_data, aes(x = PC1, y = PC2, color = cluster, label = country)) +
  geom_point(size = 2) +
  geom_text_repel(size = 3, max.overlaps = 20) +
  stat_ellipse(aes(group = cluster), type = "norm", level = 0.95) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey") +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey") +
  labs(title = "PCA Cluster Plot with Country Labels and Ellipses",
       x = "Principal Component 1",
       y = "Principal Component 2") +
  theme_minimal() +
  theme(legend.title = element_text(size = 10), legend.text = element_text(size = 8))

```

```{r}
# Load necessary libraries
library(factoextra)
library(ggplot2)
library(ggrepel)

# Perform PCA
#pc <- prcomp(socialism_df, center = TRUE, scale. = TRUE)

# Perform K-means clustering
set.seed(123) # for reproducibility
optimal_clusters <- 4  # Change this to 4 clusters
kmeans_socialism <- kmeans(pc$x[, 1:8], centers = optimal_clusters, nstart = 25)

# Create a data frame for visualization
pca_data <- data.frame(pc$x[, 1:2], cluster = as.factor(kmeans_socialism$cluster), country = rownames(socialism_df))

# Visualize clusters with country labels and ellipses
ggplot(pca_data, aes(x = PC1, y = PC2, color = cluster, label = country)) +
  geom_point(size = 2) +
  geom_text_repel(size = 3, max.overlaps = 20) +
  stat_ellipse(aes(group = cluster), type = "norm", level = 0.95) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey", size = 1) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey", size = 1) +
  labs(title = "PCA Cluster Plot with Country Labels and Ellipses",
       x = "Principal Component 1",
       y = "Principal Component 2") +
  theme_minimal() +
  theme(panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(),
        axis.line = element_line(color = "black", size = 0.5),
        axis.ticks = element_line(color = "black", size = 0.5),
        legend.title = element_text(size = 10), 
        legend.text = element_text(size = 8))

```

# Hierarchical Clustering

```{r}
# Compute the distance matrix using the first 8 principal components
distance_matrix <- dist(pc$x[, 1:8])
# Perform hierarchical clustering using Ward's method
hc <- hclust(distance_matrix, method = "ward.D2")

```

```{r}
# Plot the dendrogram
plot(hc, labels = rownames(socialism_df), main = "Dendrogram of Hierarchical Clustering", sub = "", xlab = "", ylab = "Height")

```

```{r}
# Load necessary library
library(cluster)

# Compute the silhouette width for different numbers of clusters
sil_width <- numeric(10)
for (k in 2:10) {
  pam_fit <- pam(distance_matrix, diss = TRUE, k = k)
  sil_width[k] <- pam_fit$silinfo$avg.width
}

# Plot silhouette width
plot(1:10, sil_width, type = "b", xlab = "Number of clusters", ylab = "Silhouette Width", main = "Silhouette Method")

```

```{r}
# Cut the dendrogram to form clusters
hc_clusters <- cutree(hc, k = 3)

# Create a data frame for visualization and analysis
pca_data <- data.frame(pc$x[, 1:2], hc_cluster = as.factor(hc_clusters), country = rownames(socialism_df))

# Visualize hierarchical clusters with PCA scores
ggplot(pca_data, aes(x = PC1, y = PC2, color = hc_cluster, label = country)) +
  geom_point(size = 2) +
  geom_text_repel(size = 3, max.overlaps = 20) +
  stat_ellipse(aes(group = hc_cluster), type = "norm", level = 0.95) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey", size = 1) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey", size = 1) +
  labs(title = "PCA Cluster Plot with Hierarchical Clusters",
       x = "Principal Component 1",
       y = "Principal Component 2") +
  theme_minimal() +
  theme(panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(),
        axis.line = element_line(color = "black", size = 0.5),
        axis.ticks = element_line(color = "black", size = 0.5),
        legend.title = element_text(size = 10), 
        legend.text = element_text(size = 8))

```

Applying different Methods

```{r}
# Perform hierarchical clustering using Complete Method 
hc_2 <- hclust(distance_matrix, method = "complete")

```

```{r}
# Plot the dendrogram
plot(hc_2, labels = rownames(socialism_df), main = "Dendrogram of Hierarchical Clustering", sub = "", xlab = "", ylab = "Height")
```

If we draw

```{r}
# Cut the dendrogram to form clusters
hc_clusters_2 <- cutree(hc_2, k = 2)

# Create a data frame for visualization and analysis
pca_data <- data.frame(pc$x[, 1:2], hc_cluster_2 = as.factor(hc_clusters_2), country = rownames(socialism_df))

# Visualize hierarchical clusters with PCA scores
ggplot(pca_data, aes(x = PC1, y = PC2, color = hc_cluster_2, label = country)) +
  geom_point(size = 2) +
  geom_text_repel(size = 3, max.overlaps = 20) +
  stat_ellipse(aes(group = hc_cluster_2), type = "norm", level = 0.95) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey", size = 1) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey", size = 1) +
  labs(title = "PCA Cluster Plot with Hierarchical Clusters",
       x = "Principal Component 1",
       y = "Principal Component 2") +
  theme_minimal() +
  theme(panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(),
        axis.line = element_line(color = "black", size = 0.5),
        axis.ticks = element_line(color = "black", size = 0.5),
        legend.title = element_text(size = 10), 
        legend.text = element_text(size = 8))
```

# Top socialists country in the world

in python
